
---
title: "Intro to Generalized Mixed Effects Model in R"
external_link: ''
date: 'Updated at: `r Sys.Date()`'
links:
- icon: twitter
  icon_pack: fab
  name: Follow
  url: https://twitter.com/hanhao23
slides: null
summary: An example of generalized mixed effects model (GLMM) in with lmer package.
tags:
- Repeated Measures
- Mixed Effects Model
- R Stuff
- Stats
- Psychometrics
- Academic
image:
  focal_point: Smart
url_code: ''
url_pdf: ''
url_slides: ''
url_video: ''
---

# Overview

The goal of this document is to introduce applications of R for Generalized Mixed Effects Modeling.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```

## Package and Data Preparation

```{r cars}
library(psych) # For descriptives and ANOVA, and else
library(ez) # For ANOVA
library(tidyverse) # This is a collection of packages for data wrangling and visualizing
library(Rmisc)
library(reshape2) # For reorganizing data
#library(lsr)
library(lme4) # For generalized linear mixed effect model
library(lmerTest) # For p values in generalized linear mixed effect model
#library(emmeans)
#library(dplyr)
#library(forcats)
library(DescTools)
#library(SuppDists)
library(effsize)
library(ggpubr)
#library(MVN)
library(r2glmm)

dat <- read.csv("Demo Data Das.csv") # Read in the csv data
```

## Data forging


```{r forging}

# A quick look at the first several rows of the data. This is the wide format in which each row contains all information from one individual.
head(dat)

# This function melt a wide-format data into long-formant in which each row contains information from one trial.
dat2 <- melt(dat, measure.vars = c("D1S","D2S","D3S", "D1D","D2D","D3D"), variable.name = "Condition", value.name = "Score")

# Spliting the string variable "condition" into two seperate (repeated measure) variables
dat3 <- separate(dat2, Condition, sep = 2, into = c("Factor1","Factor2"), remove = TRUE)

# I recoded the ID variable to a factor (for the ANOVA analyses, otherwise R will treat it as a DV)
dat3$X <- as.factor(dat3$X)
# Factor 1 has 3 levels and I took out the 3rd level (otherwise I will have to have 2 dummy-coded variables for Factor1 in regression).
dat3Final <-subset(dat3, Factor1 != "D3")

# Now the current data are formatted as a 'perfect' long format
head(dat3)


```

## ANOVA with "ezANOVA" Package

This package gives 3 options for us to calculate Sums of Squares, and the following note is copied directly from their documentation:  

>Numeric value (either 1, 2 or 3) specifying the Sums of Squares “type” to employ when data are unbalanced (eg. when group sizes differ). type = 2 is the default because this will yield identical ANOVA results as type = 1 when data are balanced but type = 2 will additionally yield various assumption tests where appropriate. When data are unbalanced, users are warned that they should give special consideration to the value of type. type=3 will emulate the approach taken by popular commercial statistics packages like SAS and SPSS, but users are warned that this approach is not without criticism.
    
```{r ezANOVA}

# Type 1
ezANOVA(dat3Final, dv = .(Score), wid = .(X), within = .(Factor1, Factor2), type = 1, return_aov = TRUE, detailed = TRUE)

# Type 3
ezANOVA(dat3Final, dv = .(Score), wid = .(X), within = .(Factor1, Factor2), type = 3, return_aov = TRUE, detailed = TRUE)


```

## ANOVA with "aov" Function

```{r anova}
AnovaModel <- aov(Score ~ Factor1*Factor2 + Error(X/(Factor1*Factor2)), data = dat3)
summary(AnovaModel)
# Effect sizes
EtaSq(AnovaModel, type = 1)

```

## ANOVA Plot

This is just a quick visualization of the condition differences.

```{r ANOVAViz}
DescribeSummary <- summarySE(dat3Final, measurevar = "Score", groupvars = c("Factor1","Factor2"))

pd = position_dodge(0.9)

ggplot(DescribeSummary, aes(x=Factor1, y=Score, fill=Factor2)) + 
  geom_errorbar(aes(ymin=Score-se, 
                    ymax=Score+se), 
                width=.2, size=1, position=pd) +
  geom_bar(position = "dodge", stat = "identity", alpha = 0.7) +
  coord_cartesian(ylim=c(2,9))+
  theme_classic() +
  scale_fill_grey(start = .1, end = .8) +
  theme(
    axis.title.y = element_text(vjust= 1.8),
    axis.title.x = element_text(vjust= -0.5),
    axis.title = element_text(face = "bold"))

```


## Generalized Linear Mixed Effect Model

I just attached the scripts that I used in my own project, but I also just started to use GLMM so there is still a huge lot I am not 100% clear about the analysis.  
Here I dummy-coded the two factors, and specify the random intercept without considering any individual level effect of the fectors (nothing except (1|X) in the "random term" in the formula).  

This is some data I fictioned so here it seems the fitted model "lmmodel1" is singular: there might be too few variance in at least one effect, or it could also be a miss specification of the model.

```{r GLMM}
# Dummy coding
dat3Final$F1Dummy <- dummy(dat3Final$Factor1,"D2")
dat3Final$F2Dummy <- dummy(dat3Final$Factor2,"S")

# Model specification and estimation
lmmodel1 <- lmer(Score ~ WMC*F1Dummy*F2Dummy + (1|X), data = dat3Final, REML = FALSE)
summary(lmmodel1)

# Looking at the group effects
anova(lmmodel1, type = 3)

# Effect sizes. I googled and found this R2beta people report but I am still trying to understand what it really means.
r2beta(model = lmmodel1, partial = T)

```

## GLMM Visualization

I use these codes to visualize my GLMM data. It should be of help to break the conditions down and visualize the correlations beween the continous predictor and the outcome (performance) by unique conditions.

```{r GLMMViz}
split_plot <- ggplot(aes(WMC, Score), data = dat3Final) + 
  geom_point() + 
  stat_smooth(method = "lm", col = "red", size = 2, alpha = 0.3) +
  facet_wrap(~ Factor1*Factor2) + 
  theme_classic2() + 
  xlab("WMC") + 
  ylab("Test score")
split_plot

```


